An R Markdown document converted from "01_Clustering.ipynb

### 2023-07-29 01_Clustering.Rmd
Add files via upload

---
title: An R Markdown document converted from "GuidelineValidation.ipynb"
output: html_document
---

@@

```{r}
suppressMessages({
    library(tibble)
    library(readr)
    library(dplyr)
    library(tidyr)
    library(purrr)
    library(uwot)
    library(igraph)
    library(mvtnorm)
    library(survival)
})
```

@@

## Disease prevalence

The next step is to check whether the prevalence of certain diseases differ in the clusters found, which would give us an initial idea of the clinical relevance of these subgroups. To do that we expect a table where each column represents a disease, coded as 1 or 0 depending on the presence or absence of the disease for each individual ***included in the clustering analysis***:

```{r}
disease_dat <- inner_join(read_tsv("../data/selected_dx.tsv", show_col_types = FALSE),
                          read_tsv("../data/anycadat.tsv", show_col_types = FALSE),
                          by = "eid") %>%
    ## Only people who were included in the clustering analysis
    filter(eid %in% recoded_dat$eid)
head(disease_dat)
```

Diseases in this table are commonly associated with obesity; the list includes:

```{r}
colnames(disease_dat)[-1]
```

***Note:*** We don't expect to have all diagnoses available in all validation cohorts, but the more the better!

The first thing we need is a table summarising the sex-specific prevalence of each disease:

```{r}
overall_diseasesum <- get_diseasesummary(disease_dat, strat_dat)
```

```{r}
head(overall_diseasesum)
```

Then, we will need a table summarising the prevalence within each cluster, which is again calculated using the cluster probabilites as weights:

```{r}
cluster_diseasesum <- get_diseasesummary_clusters(disease_dat, cluster_results)
```

```{r}
head(cluster_diseasesum)
```

Some of the prevalences may be affected by subtle residual differences in the covariates we adjusted for before clustering (i.e., age, smoking and most importantly, BMI), and also certain medications. To control for that:

- We will treat 'cluster 0', the cluster characterized by all biomarkers being aligned with BMI, as the reference cluster.
- For each disease, we will calculate the increase in disease OR that corresponds to a unit increase in cluster allocation probability of every cluster relative to 'cluster 0', while all other covariates remain equal.

To do that we need an additional table containing medication data, which should look like this:

```{r}
meds_dat <- read_tsv("../data/selected_meds.tsv", show_col_types = FALSE)
head(meds_dat)
```

We defined people receiving medications using the following ATC codes:

|Medication group|ATC codes|
|:-:|:-:|
|Insulin|A10A|
|Other antidiabetic|A10B|
|Antihypertensives|C01|
||C02|
||C03|
||C07|
||C08|
||C09|
|Lipid lowering|C10|

The function to run an adjusted analysis is the following:

```{r}
cluster_adjdiseasesum <- get_adjdiseasesummary_clusters(cluster_results, disease_dat, meds_dat, strat_dat)
```

```{r}
head(cluster_adjdiseasesum)
```

These latter estimates represent the change in the logarithm of the OR for a disease per each unit change in the logarithm of cluster probability. The rationale for using this transformation is to back-transform the estimates so that they represent the change in the OR for a certain percent increase/decrease in probability.

As a complementary analysis, we would calculate these OR by allocating individuals to a certain cluster based on different thresholds. The function for that is the following:

```{r}
cluster_adjdiseasesum_thresh <- get_adjdiseasesummary_clustersthresh(cluster_results, disease_dat, meds_dat, strat_dat, thresh = c(.6, .7, .8))
```

```{r}
head(cluster_adjdiseasesum_thresh)
```

These estimates represent the change in the logarithm of the OR for change in cluster membership relative to cluster 0 - the concordant cluster.

## MACE incidence - UK Biobank, Rotterdam (GHS?)

Next, we will calculate the risk of future disease given the probability of cluster allocation. The table we expect to able to run this part looks like the following:

```{r}
survmacedat <- read_tsv("../data/survmacedat.tsv", show_col_types = FALSE)
head(survmacedat)
```

The first thing we need, as done before, is a sex-specific summary of the data available:

```{r}
macesum <- get_incidencesummary(survmacedat, strat_dat)
```

```{r}
macesum
```

And again, we would run similar cluster-specific analyses as done for prevalence:

- Crude incidence per cluster:

```{r}
cluster_macesum <- get_incidencesummary_clusters(survmacedat, cluster_results)
```

```{r}
cluster_macesum
```

- Unadjusted and adjusted survival models with cluster probabilities as predictors:

```{r}
cluster_adjmacesum <- get_adjincidencesummary_clusters(cluster_results, survmacedat, meds_dat, strat_dat)
```

```{r}
head(cluster_adjmacesum)
```

- Unadjusted and adjusted survival models with hard cluster allocation at different thresholds as predictor:

```{r}
cluster_adjmacesum_thresh <- get_adjincidencesummary_clustersthresh(cluster_results, survmacedat, meds_dat, strat_dat, thresh = c(.6, .7, .8))
```

```{r}
head(cluster_adjmacesum_thresh)
```

We want to know whether a simple survival model using all biomarkers is able to capture the risk we observe in the clusters. To do that we first construct this simple survival model:

```{r}
globalmod_mace <- global_survivalmodel(strat_dat, survmacedat)
```

The result is a sex specific survival model:

```{r}
lapply(globalmod_mace, class)
```

Each model includes all variables, e.g.:

```{r}
summary(globalmod_mace$Female)$coefficients
```

We then extract the probability assigned by these models to each individual to be free of MACE events after 5 years of follow-up, and then we assess both the overall and the cluster-specific predictive performance. For the first we have the following function:

```{r}
global_survmetrics_mace <- global_survmetrics(globalmod_mace, strat_dat, survmacedat)
```

This function returns for each sex a list containing 2 elements:
- A table containing multiple thresholds of survival probabilities, and their corresponding number of true/false positives and true/false negatives, which we can use to calculate ROC curves.
- A calculation of the variance parameters around the ROC AUC estimate, which we will use to formally compare 2 ROC AUCs using Delong's method.

For example, these is how this object looks like for females:

```{r}
str(global_survmetrics_mace$Female)
```

We will also calculate these same estimates for each cluster. We will do this by weighting individuals by their cluster probabilities:

```{r}
cluster_survmetrics_mace <- cluster_survmetrics(globalmod_mace, strat_dat, survmacedat, cluster_results)
```

The result is similar as before, for each cluster identified:

```{r}
str(cluster_survmetrics_mace$Female$cluster_1)
```

## Diabetes risk - UK Biobank, Rotterdam (GHS?)

Given the strong link between BMI and T2D, we will also test whether one of the clusters identified is associated with higher or lower risk of developing diabetes over time. The data we need looks just like the one we used above for MACE:

```{r}
survdmdat <- read_tsv("../data/survdmdat.tsv", show_col_types = FALSE)
head(survdmdat)
```

We will then apply the same functions as we did for MACE:

- Sex-specific summary of the data available:

```{r}
timetodmsum <- get_incidencesummary(survdmdat, strat_dat)
```

```{r}
timetodmsum
```

- Crude incidence per cluster:

```{r}
cluster_timetodmsum <- get_incidencesummary_clusters(survdmdat, cluster_results)
```

```{r}
head(cluster_timetodmsum)
```

- Unadjusted and adjusted survival models with cluster probabilities as predictors (here ignoring medication):

```{r}
cluster_adjtimetodmsum <- get_adjincidencesummary_clusters(cluster_results, survdmdat, meds_dat[,"eid"], strat_dat)
```

```{r}
head(cluster_adjtimetodmsum)
```

- Unadjusted and adjusted survival models with hard cluster allocation at different thresholds as predictor:

```{r}
cluster_adjtimetodmsum_thresh <- get_adjincidencesummary_clustersthresh(cluster_results, survdmdat, meds_dat[,"eid"], strat_dat, thresh = c(.6, .7, .8))
```

```{r}
head(cluster_adjtimetodmsum_thresh)
```

- Overall survival model:

```{r}
globalmod_timetodm <- global_survivalmodel(strat_dat, survdmdat)
```

```{r}
lapply(globalmod_timetodm, class)
```

- Overall predictive performance at 5 years:

```{r}
global_survmetrics_timetodm <- global_survmetrics(globalmod_timetodm, strat_dat, survdmdat)
```

- Cluster-specific predictive performance at 5 years, weighted by probabilities:

```{r}
cluster_survmetrics_timetodm <- cluster_survmetrics(globalmod_timetodm, strat_dat, survdmdat, cluster_results)
```

@@

```{r}
result_file <- list(
    General_descriptives = gendesc_tab,
    BMI_coefficients = bmicoefs_tab,
    Cluster_results = cluster_results,
    Cluster_summary = clustersummary,
    Cluster_descriptives = cluster_descriptives,
    Cluster_BMI_coefficients = clusbmicoefs,
    Disease_summary = overall_diseasesum,
    Cluster_Disease_summary = cluster_diseasesum,
    Cluster_Disease_summary_Adjusted = cluster_adjdiseasesum,
    Cluster_Disease_summary_Threshold = cluster_adjdiseasesum_thresh,
    MACE_summary = macesum,
    Cluster_MACE_summary = cluster_macesum,
    Cluster_MACE_summary_Adjusted = cluster_adjmacesum,
    Cluster_MACE_summary_Threshold = cluster_adjmacesum_thresh,
    Global_SurvMetrics_MACE = global_survmetrics_mace,
    Cluster_SurvMetrics_MACE = cluster_survmetrics_mace,
    TimetoDM_summary = timetodmsum,
    Cluster_TimetoDM_summary = cluster_timetodmsum,
    Cluster_TimetoDM_Adjusted = cluster_adjtimetodmsum,
    Cluster_TimetoDM_Threshold = cluster_adjtimetodmsum_thresh,
    Global_SurvMetrics_TimetoDM = global_survmetrics_timetodm,
    Cluster_SurvMetrics_TimetoDM = cluster_survmetrics_timetodm
)
```

@@

The results shown do not contain any individual level data. We require that you upload this 'result_file.RData' file to the Teams folder in SOPHIA, which would be located here:

@@

>>>>

---
title: An R Markdown document converted from "01_Clustering.ipynb"
output: html_document
---

@@

```{r}
suppressMessages({
    library(tibble)
    library(readr)
    library(dplyr)
    library(tidyr)
    library(purrr)
    library(uwot)
    library(igraph)
    library(mvtnorm)
})
```

@@

@@

```{r}
result_file <- list(
    General_descriptives = gendesc_tab,
    BMI_coefficients = bmicoefs_tab,
    Cluster_results = cluster_results,
    Cluster_summary = clustersummary,
    Cluster_descriptives = cluster_descriptives,
    Cluster_BMI_coefficients = clusbmicoefs
)
```

@@

We require that you upload the 'result_file.RData' file, which does not contain any individual level data, to the Teams folder in SOPHIA, which would be located here:

@@

## Saving discovery cohort used

```{r}
save(strat_dat, file = "../data/ukb/strat_dat.RData")
```
